---
title: "DATA 606 - Final Project Proposal"
author: "Adam Douglas"
date: "10/30/2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(forcats)
```

### Data Preparation

The data comes from the American National Election Studies (ANES), a survey conducted by is a collaboration between Stanford University and the University of Michigan, with funding by the National Science Foundation.[^1]

```{r load, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
# Load data via a pipe-delimited file, using character type for all columns
rawData <- read_delim("anes_timeseries_2016_rawdata.txt","|",
                      col_types= paste(rep("c",1290),sep="",collapse=""))

# Empty data frame
survey <- data.frame(ID = as.character(rawData$V160001))

# Recode some initial variables
survey$surveyMethod <- fct_recode(rawData$V160501,
                                  "Face-To-Face" = "1", "Web" = "2")

survey$surveyType <- fct_recode(rawData$V160502,
                                "Only Pre" = "0", "Pre and Post" = "1")

survey$sex <- fct_recode(rawData$V161002,
                         "Male" = " 1", "Female" = " 2", NULL = "-1")

survey$payAttPol <- fct_recode(rawData$V161003, 
                               "Always" = "1", "Most of the time" = "2", 
                               "About half of the time" = "3", 
                               "Some of the time" = "4", "Never" = "5")

survey$votePrim <- fct_recode(rawData$V161021, 
                              "Yes" = " 1", "No" = " 2", 
                              NULL = "-8", NULL = "-9")

survey$earlyVote <- fct_recode(rawData$V161022, 
                               "Yes" = " 1", "No" = " 2",
                               NULL = "-1", NULL = "-9")

survey$clinton <- as.numeric(rawData$V161086)
survey$clinton[survey$clinton < 0] <- NA

survey$trump <- as.numeric(rawData$V161087)
survey$trump[survey$trump < 0] <- NA

survey <- survey %>% mutate(bothCandidates = trump - clinton)
```

### Research Question

My research question is: does a correlation exist between a person's exposure to certain news media outlets (web and traditional) and their stated attitudes towards the presidential candidates in the 2016 election?

### Cases

Each case is a person surveyed. There are 4271 observations in the data set.

### Data Collection

Data was collected by survey, either in-person or via the web. The larger data set has people surveyed both before and after the election, but we will look at questions answered just prior to the election.

### Type of Study

This is an observational study

### Data Source

Data was retreived from the [ANES website](https://electionstudies.org/data-center/)

The [American National Election Studies](www.electionstudies.org). These materials are based on work supported by the National Science Foundation under grant numbers SES 1444721, 2014-2017, the University of Michigan, and Stanford University. Any opinions, findings and conclusions, or recommendations expressed in these materials are those of the author(s) and do not necessarily reflect the views of the funding organizations.

### Response

The response variable that we will be using is numeric and is the respondent's rating of then-candidate Donald Trump minus their rating of Hillary Clinton.

### Explanatory

The explanatory variables are the respondent's use of various media (web, print, TV) and frequency (where given).

## Relevant summary statistics

```{r}
survey %>% filter(!is.na(clinton), !is.na(trump)) %>% summary(.)
```

```{r}
# Tidy variables for some summary statistics
ratings <- survey %>% gather(key = "candidate", value = "rating", 
                             clinton, trump, bothCandidates) %>% 
  filter(earlyVote == "No", votePrim == "No") %>%
  select(payAttPol, candidate, rating)

# Box plot of individual candidate ratings
ratings %>% filter(!is.na(rating)) %>% 
  ggplot(aes(x=candidate, y=rating, fill=candidate)) +
  geom_boxplot(fill=c("purple","blue","red")) +
  ggtitle("Respondent Ratings of Candidates") +
  xlab("Candidate") +
  ylab("Rating") +
  scale_x_discrete(labels=c("Both Candidates","Hillary Clinton","Donald Trump"))

# Box plot of Trump ratings by political attentiveness
ratings %>% filter(candidate=="trump", !is.na(rating)) %>%
  ggplot(aes(x=payAttPol, y=rating)) +
  geom_boxplot(fill="red") +
  ggtitle("Respondent Ratings of Donald Trump","By Political Mindfulness") +
  xlab("How Often Pay Attention to Politics") +
  ylab("Rating") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Box plot of Clinton ratings by political attentiveness
ratings %>% filter(candidate=="clinton", !is.na(rating)) %>%
  ggplot(aes(x=payAttPol, y=rating)) +
  geom_boxplot(fill="blue") +
  ggtitle("Respondent Ratings of Hillary Clinton","By Political Mindfulness") +
  xlab("How Often Pay Attention to Politics") +
  ylab("Rating") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```




[^1]: The American National Election Studies (www.electionstudies.org).